---
title: "Comparing Different Multilevel Bootstrapping Methods"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Different Multilevel Bootstrapping Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Downloading the Data

Here I use the example data in the book by Hox (2010, p. 17). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
comma <- function(x) format(x, digits = 2, big.mark = ",")
```


```{r}
library(tidyverse)
library(haven)
library(lme4)
library(boot)
library(bootmlm)
popdata <- haven::read_dta(
  "https://stats.idre.ucla.edu/stat/stata/examples/mlm_ma_hox/popular.dta")
```

For the interest of time, I will select only 30 schools and a few students in 
each school:

```{r}
set.seed(85957)
pop_sub <- popdata %>% filter(school %in% sample(unique(school), 30)) %>%
  group_by(school) %>% sample_frac(size = .25) %>% ungroup()
```

To illustrate doing bootstrapping to obtain standard error (*SE*) and confidence
interval (CI) for the intraclass correlation (ICC) of the variable `popular`, we
first fit an intercept only model:

```{r}
m0 <- lmer(popular ~ (1 | school), data = pop_sub)
summary(m0)
```

The intraclass correlation is defined as
$$\rho = \frac{\tau}{\tau + \sigma^2} = \frac{1}{1 + \sigma^2 / \tau} 
  = \frac{1}{1 + \theta^{-2}},$$
where $\theta = \sqrt{\tau} / \sigma$ is the relative cholesky factor for the
random intercept term used in `lme4`. Therefore, we can estimate the ICC as:

```{r}
(icc0 <- 1 / (1 + getME(m0, "theta")^(-2)))
```

So the ICC is quite large for this data set. However, it is important to also
quantify the uncertainty of a point estimate. Although there are analytic 
methods to obtain *SE* and CI for ICC, a reliable alternative is to do 
bootstrapping. We first define the function for computing the test statistic:

```{r}
icc <- function(x) 1 / (1 + x@theta^(-2))
icc(m0)
```

Note that I used the `@` operator for faster extraction. With the `bootmlm`
package we can perform various bootstrap methods using the `bootstrap_mer()`
function.

## Parametric Bootstrap

We can run parametric bootstrap, which essentially call the `lme4::bootMer()` 
function. It's usually recommended to have a large number of bootstrap samples
($R$), especially for CIs with higher confidence levels. For illustrative 
purpose I will use $R = 999$, but in general 1,999 or more is recommended

```{r}
system.time(
  boo_par <- bootstrap_mer(m0, icc, nsim = 999L, type = "parametric"))
boo_par
```

As you can see, the *SE* for the ICC is estimated to be `r sd(boo_par$t)` with
parametric bootstrap. 

### Confidence interval

With parametric bootstrap there are three ways to construct confidence 
intervals via the `boot.ci()` function from the `boot` package: normal, 
basic, and percentile. We can use the following function:

```{r}
boo_par_ci <- boot::boot.ci(boo_par, type = c("norm", "basic", "perc"), 
                            index = 1L)
boo_par_ci
```


## Residual Bootstraps

Whereas parametric bootstrap resamples from independent normal distributions, 
residual bootstrap samples the residuals. Therefore, residual bootstrap is 
expected to be more robust to non-normality. `bootmlm` implements three methods
for residual bootstrap: differentially reflated residual bootstrap, 
Carpenter-Goldstein-Rashbash's residual bootstrap (CGR; Carpenter et al., 2003),
and transformational residual bootstrap by van der Leeden, Meijer, and Busin 
(2008). They are all motivated by the fact that the residuals, generally 
empirical bayes estimates (denoted as $\tilde u$ and $\tilde e$), are shrinkage
estimates and have sampling variabilities much smaller than the population
random effects, $u$ and $e$.

The first residual bootstrap rescale $\tilde u$ and $\tilde e$ so that their
sampling variabilities match those of $u$ and $e$ as implied by the model 
estimates. This can be obtained by 

```{r}
system.time(
  boo_res <- bootstrap_mer(m0, icc, nsim = 999L, type = "residual"))
boo_res
```

As you can see, the *SE* for the ICC is estimated to be `r sd(boo_res$t)` with
residual bootstrap. 

The second method, CGR, rescale the sample covariance matrix of the 
*realized values* of the residuals to match the model-implied variance 
components. This can be obtained by 

```{r}
system.time(
  boo_cgr <- bootstrap_mer(m0, icc, nsim = 999L, type = "residual_cgr"))
boo_cgr
```

The *SE* is estimated to be `r sd(boo_cgr$t)` with CGR bootstrap. 

The third method first transforms the OLS residuals, $\hat r_{ij} = y_{ij} -
\boldsymbol{x}_{ij} \hat{\boldsymbol{\beta}}$, by the inverse of cholesky
factor, $\boldsymbol{L}$, of the model-implied covariance matrix of
$\boldsymbol{y}$, $\hat{\boldsymbol{V}}$, so that theoretically 
$\boldsymbol{L}^{-1} (\boldsymbol{y} - \boldsymbol{X \beta})$ should be 
independent and identically distributed. However, as the true sampling variance
of $\hat r_{ij}$ is not $\boldsymbol{V}$, I also provide the option
`corrected_trans = TRUE` to do the transformation using the theoretically
sampling variability of $\hat r_{ij}$.

```{r}
# Transformation according to V
system.time(
  boo_tra <- bootstrap_mer(m0, icc, nsim = 999L, type = "residual_trans"))
boo_tra
# Transformation according to the sampling variance of r
system.time(
  boo_trac <- bootstrap_mer(m0, icc, nsim = 999L, type = "residual_trans", 
                            corrected_trans = TRUE))
boo_trac
```

The *SE* is estimated to be `r sd(boo_tra$t)` and `r sd(boo_trac$t)` with
and without corrections with the transformational residual bootstrap. 

### Confidence interval

With residual bootstrap methods there are four ways to construct confidence 
intervals via the `boot.ci()` function from the `boot` package, with the 
addition of the bias-corrected and accelarted bootstrap (BCa). We can use the
following function:

```{r}
# First need to compute the influence values
inf_val <- empinf_mer(m0, icc, index = 1)
# Residual bootstrap
boo_res_ci <- boot::boot.ci(boo_res, type = c("norm", "basic", "perc", "bca"), 
                            index = 1L, L = inf_val)
boo_res_ci
# CGR
boo_cgr_ci <- boot::boot.ci(boo_cgr, type = c("norm", "basic", "perc", "bca"), 
                            index = 1L, L = inf_val)
boo_cgr_ci
# Transformational (no correction)
boo_tra_ci <- boot::boot.ci(boo_tra, type = c("norm", "basic", "perc", "bca"), 
                            index = 1L, L = inf_val)
boo_tra_ci
# Transformational (with correction)
boo_trac_ci <- boot::boot.ci(boo_trac, type = c("norm", "basic", "perc", "bca"), 
                             index = 1L, L = inf_val)
boo_trac_ci
```

## Random Effect Block Bootstrap

```{r}
system.time(
  boo_reb <- bootstrap_mer(m0, icc, nsim = 999L, type = "reb"))
boo_reb
system.time(
  boo_rebs <- bootstrap_mer(m0, icc, nsim = 999L, type = "reb", 
                            reb_scale = TRUE))
boo_rebs
```

### Confidence interval

```{r}
# Only sampling clusters
boo_reb_ci <- boot::boot.ci(boo_reb, type = c("norm", "basic", "perc", "bca"), 
                            index = 1L, L = inf_val)
boo_reb_ci
# Transformational (with correction)
boo_rebs_ci <- boot::boot.ci(boo_rebs, type = c("norm", "basic", "perc", "bca"), 
                             index = 1L, L = inf_val)
boo_rebs_ci
```

## Case Bootstrap

With case bootstrap, the observed *cases* are sampled with replacement. 
However, because of the multilevel structure, we need to resample the 
clusters. Optionally, we can then resample the cases within each cluster
(using the `lv1_resample = TRUE` argument). Unlike the parametric and 
residual bootstrap methods, currently `bootmlm` only support the case bootstrap 
with two levels. 

```{r}
system.time(
  boo_cas <- bootstrap_mer(m0, icc, nsim = 999L, type = "case"))
boo_cas
system.time(
  boo_cas1 <- bootstrap_mer(m0, icc, nsim = 999L, type = "case", 
                            lv1_resample = TRUE))
boo_cas1
```

The *SE* for the ICC is estimated to be `r sd(boo_cas$t)` (only sampling 
clusters) and `r sd(boo_cas1$t)` (sampling also cases) with case bootstrap. 

### Confidence interval

With case bootstrap the supported CIs are: normal, basic, and percentile, and 
BCa. We can use the following function:

```{r}
# Only sampling clusters
boo_cas_ci <- boot::boot.ci(boo_cas, type = c("norm", "basic", "perc", "bca"), 
                            index = 1L, L = inf_val)
boo_cas_ci
# Transformational (with correction)
boo_cas1_ci <- boot::boot.ci(boo_cas1, type = c("norm", "basic", "perc", "bca"), 
                             index = 1L, L = inf_val)
boo_cas1_ci
```

## Summary

```{r}
boo_names <- c("parametric", "residual", "cgr", "trans", 
               "trans (cor)", "REB", "REB (scaled)", 
               "case (cluster)", "case (c + i)")
boo_lst <- list(boo_par, boo_res, boo_cgr, boo_tra, boo_trac, 
                boo_reb, boo_rebs, boo_cas, boo_cas1)
boo_ci_lst <- list(boo_par_ci, boo_res_ci, boo_cgr_ci, boo_tra_ci, 
                   boo_trac_ci, boo_reb_ci, boo_rebs_ci, boo_cas_ci, 
                   boo_cas1_ci)
get_ci <- function(boo_ci, type) {
  paste0("(", paste(comma(tail(boo_ci[[type]][1, ], 2L)), collapse = ", "), ")")
}

data_frame(boot_type = boo_names, boo = boo_lst, boo_ci = boo_ci_lst) %>%
  mutate(sd = map_chr(boo, ~ comma(sd(.x$t))), 
         normal = map_chr(boo_ci, ~ get_ci(.x, "normal")), 
         basic = map_chr(boo_ci, ~ get_ci(.x, "basic")), 
         percentile = map_chr(boo_ci, ~ get_ci(.x, "percent")), 
         bca = map_chr(boo_ci, ~ get_ci(.x, "bca"))) %>%
  select(-boo, -boo_ci)
```


<!-- Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format: -->

<!-- - Never uses retina figures -->
<!-- - Has a smaller default figure size -->
<!-- - Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style -->

<!-- ## Vignette Info -->

<!-- Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette. -->

<!-- ## Styles -->

<!-- The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows: -->

<!--     output:  -->
<!--       rmarkdown::html_vignette: -->
<!--         css: mystyles.css -->

<!-- ## Figures -->

<!-- The figure sizes have been customised so that you can easily put two images side-by-side.  -->

<!-- ```{r, fig.show='hold'} -->
<!-- plot(1:10) -->
<!-- plot(10:1) -->
<!-- ``` -->

<!-- You can enable figure captions by `fig_caption: yes` in YAML: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         fig_caption: yes -->

<!-- Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**. -->

<!-- ## More Examples -->

<!-- You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`. -->

<!-- ```{r, echo=FALSE, results='asis'} -->
<!-- knitr::kable(head(mtcars, 10)) -->
<!-- ``` -->

<!-- Also a quote using `>`: -->

<!-- > "He who gives up [code] safety for [code] speed deserves neither." -->
<!-- ([via](https://twitter.com/hadleywickham/status/504368538874703872)) -->
