---
title: "Comparing Different Multilevel Bootstrapping Methods"
author: "Mark Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Different Multilevel Bootstrapping Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Downloading the Data

Here I use the example data in the book by Hox (2010, p. 17). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
comma <- function(x) format(x, digits = 2, big.mark = ",")
```


```{r}
library(tidyverse)
library(haven)
library(lme4)
library(boot)
library(bootmlm)
popdata <- haven::read_dta(
  "https://stats.idre.ucla.edu/stat/stata/examples/mlm_ma_hox/popular.dta")
```

For the interest of time, I will select only 20 schools and a few students in 
each school:

```{r}
set.seed(85957)
pop_sub <- popdata %>% filter(school %in% sample(unique(school), 20)) %>%
  group_by(school) %>% sample_frac(size = .25) %>% ungroup()
```

To illustrate doing bootstrapping to obtain standard error (*SE*) and confidence
interval (CI) for the intraclass correlation (ICC) of the variable `popular`, we
first fit an intercept only model:

```{r}
m0 <- lmer(popular ~ (1 | school), data = pop_sub)
summary(m0)
```

The intraclass correlation is defined as
$$\rho = \frac{\tau^2}{\tau^2 + \sigma^2} = \frac{1}{1 + \sigma^2 / \tau^2} 
  = \frac{1}{1 + \theta^{-2}},$$
where $\theta = \tau / \sigma$ is the relative cholesky factor for the
random intercept term used in `lme4`. Therefore, we can estimate the ICC as:

```{r}
(icc0 <- 1 / (1 + getME(m0, "theta")^(-2)))
```

So the ICC is quite large for this data set. However, it is important to also
quantify the uncertainty of a point estimate. We can get the asymptotic 
standard error (ASE) of the ICC using the *delta method*:

```{r}
# Get the point estimate of theta:
th_est <- getME(m0, "theta")
# Get the variance of theta using the bootmlm::vcov_theta() function
th_var <- bootmlm::vcov_theta(m0)
# We can get the asymptotic variance of ICC using delta method with an 
# appropriate transformation. `x1` is the variable name needed for the first
# variable, which is theta in our case
icc_avar <- msm::deltamethod(g = ~ 1 / (1 + x1^(-2)), 
                            mean = th_est, cov = th_var)
```

So we can get an approximate symmetric CI with 

```{r}
icc0 + c(-2, 2) * sqrt(icc_avar)
```

## Bootstrap-$t$ CI

```{r}
icc <- function(x) {
  th_est <- x@theta
  est <- 1 / (1 + th_est^(-2))
  th_var <- vcov_theta(x)
  var <- msm::deltamethod(g = ~ 1 / (1 + x1^(-2)), 
                            mean = th_est, cov = th_var)
  c(est, var)
}
icc(m0)
boo <- bootstrap_mer(m0, icc, 999L, type = "case")
# Influence values for BCa
inf_val <- empinf_mer(m0, icc, index = 1)
boot::boot.ci(boo, L = inf_val)
```

